Specific taxa differentiate proximal and distal microbiota in the unprepped

### 2017-07-10 submission/manuscript.Rmd
updated discussion

The human colon is an ecosystem comprised of several different microenvironments inhabited by resident bacterial members of the microbiome. Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence what populations of microbes reside in each location. Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC). IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes (@Yamauchi2012). The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur (cite). For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms. In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010). The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients (@Yamauchi2012, @Albenberg2014, @Donaldson2015). In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically present with disease in the small intestine, ileum and proximal colon (cite). UC presents as large and highly inflammed mucosal ulcers, where as Crohn's disease lesions are often smaller and have areas of normal tissue distributed amongst the flare (cite). Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC at these locations. Because CRC is a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.

Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012, IBDREF). For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated (@Dejea2014). These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_. Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our human stool classification studies (@Baxter2016). _F. nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls (Strauss2011). Furthermore, studies of _F. nucleatum_ isolated from mucosal biopsies showed that more invasive  _F. nucleatum_ positively correlates with IBD disease level (Strauss2011). Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant (cite). Thus the molecular capabilities of these rare taxa may drive the colonic disease state. Previous studies examining the microbes associated with CRC and IBD have focused on overall community structure, diversity and membership and not lowly-abundant species. Further, because these studies often examined only shed human stool, they were unable to analyze paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location is largely undefined. Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct distal and proximal CRC tumors or IBD flares.

@@

Here we aimed to address the limitations of previous studies and effectively identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon. Our design used unprepared colonoscopy techniques to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers. To address the inherent inter-individual variation in human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location. We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes. By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.

@@

We detected _F. nucleatum_ and _P. asacharolytica_ in 8 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans @Baxter2016 and have oncogenic properties in cell culture and in mice @Sears2014. Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the distal mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous distal mucosa (@Albenberg2014). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on proximal-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (@Mima2016).  Of the 8 (40%) individuals positive for _F. nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa. The presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers (cite). Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis (@Dejea2014). Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016). Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of CRC tumors in the proximal or distal colon.

Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies. A rare report of a matched-colonoscopy study that sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (@Shobar2016). This group found that mucosa and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (@Shobar2016). Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites. Further, the differences were not great enough to overcome interpersonal differences between subjects. So, bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel. Thus our findings of specific bacteria at each site of the colon are strengthened by the lack of bowel preparation.

Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer. Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors. Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit and it performed much better than FIT alone. Further work from our lab has shown that microbiome profiling of the FIT cartridge contents sufficiently represented the stool community (@Baxter2016). One caveat of the FIT study was that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective. Given that our results showed that the stool most accurately reflects the community of the distal lumen, we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.

>>>>

The human colon is an ecosystem comprised of several different microenvironments inhabited by resident bacterial members of the microbiome. Concentrations of oxygen, water and anti-microbial peptides change along the gut axis and influence what populations of microbes reside in each location. Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. The identity of these specific microbes and communities are important for understanding the etiology of complex colon diseases such as Inflammatory Bowel Disease (IBD) and Colorectal Cancer (CRC). IBD and CRC are known to be preceded or accelerated by perturbations in gut microbes (@Yamauchi2012). The severity, symptoms, morbidity and mortality of these diseases is known to vary based upon the biogeographical location in which they occur (cite). For instance, CRC tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms. In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are sessile and form along the wall of the colon, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010). The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, in addition to variation in the previously mentioned chemical gradients (@Yamauchi2012, @Albenberg2014, @Donaldson2015). In IBD patients, disease flares in the distal colon are usually indicative of ulcerative colitis (UC), whereas Crohn's disease (CD) patients typically experience disease in the small intestine, ileum and proximal colon (cite). UC presents as large and highly inflammed mucosal ulcers, where as CD lesions are often smaller and have areas of normal tissue distributed amongst the flare (cite). Thus, given the varied physiology of the proximal-distal axis of the colon and known differences in disease patterns at these sites, symbiotic microbes and their metabolites likely vary as well, and may influence the heterogenous disease prognoses of IBD and CRC. Because CRC is a long-term complication of IBD, the distribution of microbes is important to understanding the pathophysiology of both diseases.

Several recent findings have shown that development and progression of IBD or CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012, IBDREF). For instance, comparison of the bacteria present on CRC tumors with those found on nearby healthy tissue has identified specific species that are tumor-associated (@Dejea2014). These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_. Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our human stool classification studies (@Baxter2016). _F. nucleatum_ has also been found to be elevated in the stool and biopsies of patients with IBD as compared to healthy controls (Strauss2011). Furthermore, studies of _F. nucleatum_ isolated from mucosal biopsies showed that more invasive  _F. nucleatum_ positively correlates with IBD disease level (Strauss2011). Like many intestinal pathogens, the bacteria appear to have a high-impact despite being lowly-abundant in the community (cite). The molecular capabilities of these rare taxa may drive the colonic disease state. Previous studies examining the microbes associated with CRC and IBD have focused on overall community structure, diversity and membership and not lowly-abundant species. Further, because these studies often examined only shed human stool, they were unable to analyze paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor/lesion bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Due to these limitations, the contribution of the gut microbiota to IBD and CRC disease location is largely undefined. Characterizing these communities could provide needed insight into disease etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct proximal and distal CRC tumors or IBD flares.

@@

Here we aimed to address the limitations of previous studies and effectively identify the microbes specific to the lumen and mucosa of the proximal and distal healthy human colon. Our design used an unprepared colonoscopy technique to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers. To address the inherent inter-individual variation in human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location. We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes. By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.

@@

We detected _F. nucleatum_ and _P. asacharolytica_ in 8 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans @Baxter2016 and have oncogenic properties in cell culture and in mice @Sears2014. Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the distal mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous distal mucosa (@Albenberg2014). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on proximal-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (@Mima2016).  Of the 8 (40%) individuals positive for _F. nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa. The presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers (cite). Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis (@Dejea2014). Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016). Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of IBD or CRC subtypes in the proximal or distal colon.

Specific comparisons of our findings to previously published gut biogeography studies are additionally confounded by the use of bowel preparation methods in most other studies. A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing (@Shobar2016). This group found that mucosa and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site (@Shobar2016). Bowel cleansing not only made the samples harder to distinguish, it resulted in decreases in diversity across sites. Further, the differences were not great enough to overcome interpersonal differences between subjects. So, bowel preparation clearly induces bias into the microbes recovered from sampling the lumen or mucosa of a prepared bowel. Thus our findings of specific bacteria at each site of the colon are strengthened by the lack of bowel preparation.

Microbiome-based diagnostics are increasingly being explored as non-invasive tools to survey for the development of colon cancer. Random Forest models have been used by our group and others to increase the detection sensitivity of CRC tumors. Indeed, our group found that a classification model that used microbiome data in combination with Fecal Immunohistochemical Test (FIT) results could correctly identify both carcinoma and adenoma lesions from communities of stool at exit and it performed much better than FIT alone. One caveat of the FIT study was that there was not sufficient information to test if a classification model could differentiate between proximal and distal CRC tumors based on exit stool sample alone, but we would hypothesize that would not be effective. Given that our results showed that the stool most accurately reflects the community of the distal lumen, we likely cannot use Random Forest of stool samples to diagnose any changes in the proximal or mucosal communities.
